Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-17 background.tex
+

In typical system configurations, all memory addresses seen by programs running on modern computers are \emph{virtualized}: the address observed by a running
program generally will not correspond directly to the physical location in memory, and may not even correspond to a physical location that \emph{exists} 
in the machine. Instead, these \emph{virtual} addresses are translated to \emph{physical} addresses that correspond directly to locations in RAM. On most 
modern architectures, this translation is performed through cooperation of the hardware and OS kernel: while executing an instruction that dereferences a 
(virtual) address, the CPU's \emph{memory management unit} (\textsc{MMU}) hardware performs \emph{address translation},
resulting in a physical address used to access the cache\footnote{Technically, for performance reasons most caches are indexed with parts of the virtual 
address, but tagged with the physical data addresses, so cache lookups and address translations can proceed in parallel.} and/or memory-bus.

\begin{figure}
%    \begin{framed}
%\column{.59\textwidth}    
\begin{subfigure}

@@

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.5,xscale=0.5]
%uncomment if require: \path (0,430); %set diagram left start at 0, and has height of 430

@@

% Text Node
\draw (11,14) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Virtual}};
% Text Node
\draw (570,395) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Physical}};
% Text Node
\draw (104,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 63}};
% Text Node
\draw (175,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 48}};
% Text Node
\draw (245,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 39}};
% Text Node
\draw (314,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 30}};
% Text Node
\draw (385,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 21}};
% Text Node
\draw (454,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (523,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 0}};
% Text Node
\draw (132,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Sign}};
% Text Node
\draw (195,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Offset}};
% Text Node
\draw (264,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Offset}};
% Text Node
\draw (336,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Offset}};
% Text Node
\draw (406,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Offset}};
% Text Node
\draw (472,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Page Offset}};
% Text Node
\draw (67,63) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize va}};
% Text Node
\draw (53,203) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Entry}};
% Text Node
\draw (185,189) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Entry}};
% Text Node
\draw (316,167) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Entry}};
% Text Node
\draw (449,223) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Entry}};
% Text Node
\draw (51,347) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize root}};
% Text Node
\draw (465,349) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize page}};
% Text Node
\draw (532,346) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 4K Data}};
% Text Node
\draw (156,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 51}};
% Text Node
\draw (226,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (145,350) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Base Addr}};

\end{tikzpicture}
          \caption{Accessing to the Page Referenced by L1 Entry}
        \label{fig:enter-label}
    \end{subfigure}
    \vfill
    \begin{multicols}{2}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
/*
@param entry: entry whose frame address is used to locate the a-level-lower entry in the page-tree
@return next: return a virtual pointer to the descendant tree
*/
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /*Reading page table entry justified by virtual pte-pointsto*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    /* Jump if the present bit is not zero; no need to allocate next level (* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)  (*\label{line:after_concluding_qfrac1C}*)*/
    /*Allocation path starts here (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)*/
    pte_initialize(entry);(* \label{line:call_to_pte_initializeC} *)
    /*Entry value updates:*/
    entry->pfn = nextpaddr;
    entry->present = 1; (*\label{line:install_new_entryC}*)
    /*Now we know that entry is initialized, so we satisfy the condition to access children list */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\columnbreak
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
pte_t *walkpgdir(pte_t *l4, void *va){ 
 pte_t *l4_entry = &l4[L4Offset(va)];
 pte_t *l3 = pte_nxt_table(l4_entry);
 pte_t *l3_entry = &l3[L3Offset(va)];
 pte_t *l2 = pte_nxt_table(l3_entry);
 pte_t *l2_entry = &l2[L2Offset(va)];  
 pte_t *l1 = pte_nxt_table(l2_entry);
 pte_t *l1_entry = &l1[L1Offset(va)];  
 return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{multicols}
%\end{framed}
    \vspace{-2.5em}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
    \vspace{-1em}
\end{figure}

>>>>

In typical system configurations, all memory addresses seen by programs running on modern computers are \emph{virtualized}: the address observed by a running
program generally will not correspond directly to the physical location in memory, and may not even correspond to a physical location that \emph{exists} 
in the machine. Instead, these \emph{virtual} addresses are translated to \emph{physical} addresses that correspond directly to locations in RAM. On most 
modern architectures, this translation is performed through cooperation of the hardware and OS kernel: while executing an instruction that dereferences a 
(virtual) address, the CPU's \emph{memory management unit} (\textsc{MMU}) hardware performs \emph{address translation},
resulting in a physical address used to access the cache\footnote{Technically, for performance reasons most caches are indexed with parts of the virtual 
address, but tagged with the physical data addresses, so cache lookups and address translations can proceed in parallel.} and/or memory-bus.
\setlength{\columnseprule}{4pt}
\begin{figure}\footnotesize
%    \begin{framed}
%\column{.59\textwidth}    
\begin{subfigure}\footnotesize

@@

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.75,xscale=0.75]
%uncomment if require: \path (0,427); %set diagram left start at 0, and has height of 427

@@

% Text Node
\draw (11,14) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Virtual}};
% Text Node
\draw (570,395) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Physical}};
% Text Node
\draw (104,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 63}};
% Text Node
\draw (175,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 48}};
% Text Node
\draw (245,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 39}};
% Text Node
\draw (314,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 30}};
% Text Node
\draw (385,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 21}};
% Text Node
\draw (454,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (523,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 0}};
% Text Node
\draw (132,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Sign}};
% Text Node
\draw (195,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Offset}};
% Text Node
\draw (264,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Offset}};
% Text Node
\draw (336,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Offset}};
% Text Node
\draw (406,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Offset}};
% Text Node
\draw (472,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Page Offset}};
% Text Node
\draw (67,63) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize va}};
% Text Node
\draw (48,203) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Entry}};
% Text Node
\draw (180,189) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Entry}};
% Text Node
\draw (311,167) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Entry}};
% Text Node
\draw (444,223) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Entry}};
% Text Node
\draw (51,347) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize root}};
% Text Node
\draw (465,349) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize page}};
% Text Node
\draw (529,346) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 4K Data}};
% Text Node
\draw (133,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 51}};
% Text Node
\draw (258,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (163,350) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Base Addr}};

\end{tikzpicture}
          \caption{Accessing to the Page Referenced by L1 Entry}
        \label{fig:enter-label}
    \end{subfigure}
    \vfill
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\footnotesize]
/*
@param entry: entry whose frame address is used 
              to locate the subsequent entry in the page-tree
@return next: return a virtual pointer to the descendant tree
*/
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /*Reading page table entry:
    pointed by a virtual memory address*/
  /* Jump if the present bit is not zero:
     no need to allocate next level (* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)  (*\label{line:after_concluding_qfrac1C}*)*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    /*Allocation path starts here (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)*/
    pte_initialize(entry);(* \label{line:call_to_pte_initializeC} *)
    /*Entry value updates:*/
    entry->pfn = nextpaddr;
    entry->present = 1; (*\label{line:install_new_entryC}*)
    /*Entry is initialized,
      so we satisfy the condition to access children list */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{.4\textwidth}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\footnotesize]
pte_t *walkpgdir(pte_t *l4, void *va){ 
  pte_t *l4_entry = &l4[L4Offset(va)];
  pte_t *l3 = pte_nxt_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_nxt_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_nxt_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
    \vspace{-1em}
\end{figure}
